Discrepancy, Coresets, and Sketches in Machine Learning

### 2019-01-09 komlos.tex
intro edits

\section{Introduction}
In both machine learning and in streaming and sketching problems our goal is (usually) to approximate the sums or expectations of functions.
Specifically, approximate $F = \sum_{i=1}^{n} f(x_i)$ for every $f\in \F$ where $\F$ is a family of functions and $x_i \in \mathcal X$ are either training examples or stream items.
In this paper it will often be convenient to consider an equivalent formulation.
$F(q) = \sum_{i=1}^{n} f(x_i, q)$  and $q \in \mathcal Q$ is either the model parameters (which acts as an ``index'' into $\F$) or a query from a possible set $\Q$. 
Clearly, the two formulations are identical. In this manuscript we use both interchangeably.  
For simplicity and throughout this manuscript we assume the range of $f$ is $[0,1]$.
The goal is to produce a coreset. This is a set of weights $w$ with at most $k \ll n$ non-zeros such that $\sum_{i=1}^{n}w_i f(x_i,q) := \tilde F(q)$ approximates $F(q)$.
Approximation here means that $|\tilde F(q)  - F(q)| \le \eps n$ either for all possible queries $q \in \mathcal Q$ simultaneously or for every fixed $q$ with probability at least $1-\delta$. 
There are more complicated formulations such as weak coresets which we will not touch upon in this manuscript.

Generating a concise representation $\tilde F$ for $F$ allows one to optimize over $\tilde F$ instead of $F$ which is more efficient.  Moreover, if the resulting sketches or coresets are mergable, this could be done on separate machines without the need for communication or assuming randomness in the partitioning.

For bounded functions $f$ like above, uniform sampling of $\ell = O(\log(1/\delta)/\eps^2)$ combined with a union bound over $|\mathcal Q|$ always provides a valid solution using $\ell = O(\log(|\mathcal Q|)/\eps^2)$ items. 
While $|\mathcal Q|$ is often infinite it reduces to a finite (albeit usually exponentially large) set through standard epsilon net arguments. 
%
We argue that we can offer a solution for producing smaller coresets for all of the above problems in a unified manner. 
Moreover, our solution creates streaming algorithms with fully mergeable sketches. 
The size of the coreset above appears to be intimately tied to the discrepancy properties of the associated functions.
This is captured by its Komlos complexity.

\item Kernel density estimation is a popular tool in data analysis aimed to estimate a continuous distribution with a finite set of points. Among other applications, this tool is used for outlier detection \cite{schubert2014generalized}, regression \cite{fan2018local}, and clustering \cite{rinaldo2010generalized}. A thorough survey could be found in \cite{silverman2018density}. Given a set of $n$ data points $x_1,\ldots,x_n$ and a query $q$, the density estimate of $q$ is the average $\frac{1}{n} \sum_i K(x_i,q)$ of the kernel evaluations with the data points. The task of finding a coreset for kernel density estimates consists of obtaining a set of points $z_1,\ldots,z_m$ with weights $w_1,\ldots,w_m$ for some $m \ll n$ such that for any query $q$ it holds that 
$\frac{1}{n} \sum_i K(x_i,q) = \frac{1}{m} \sum_i w_i K(z_i,q) \pm \eps$. For this task, the state-of-the-art is given by \cite{DBLP:journals/corr/abs-1802-01751}, achieving $m=\sqrt{d\log(1/\eps)}/\eps$ where $d$ is the dimension of the original data points. Their result holds for Lipchitz psd kernels. The result is constructive though it is polynomial rather than (quasi-)linear in the data size. The authors give an almost matching lower bound of $\sqrt{d}/\eps$ and pose an open question for closing the gap between the bounds.

\item Logistic Regression is an extremely popular technique for classification task. In it, we are given a set of $n$ pairs $x_i,y_i$ with $x_i \in \R^d$ and $y_i \in \{-1,1\}$. The objective is to find a regressor $q \in \R^d$ minimizing the logistic loss defined as $\E_i \log(1+\exp(x_i^\top q y_i))$. A coreset for logistic regression is a set of $m \ll n$ pairs $z_i, v_i$ in $\R^d, \{-1,1\}$ correspondingly along with weights $w_i$, with the property that for any regressor $q$, either $\E_i \log(1+\exp(x_i^\top q y_i)) = \E_i \log(1+\exp(z_i^\top q v_i)) \pm \eps$ or $\E_i \log(1+\exp(x_i^\top q y_i)) = \E_i \log(1+\exp(z_i^\top q v_i)) (1 \pm \eps)$. A recent paper \cite{DBLP:journals/corr/abs-1805-08571} provides a coreset with a multiplicative guarantee that is based on an average sensitivity property of the dataset. They provide a lower bound for the size of a multiplicative coreset showing that in general it is not possible to achieve $m \ll n$. The coreset they build is of cardinality $m \approx \mu\sqrt{nd^3}/\eps^2$ where $\mu \geq 1$ is complexity measure of the dataset.  \cite{tolochinsky2018coresets} gives a generic multiplicative coreset construction for any monotonic function with $\ell_2^2$ regularization. The dependence they get is, ignoring logarithmic factors, is $d/\eps^2$.

\section{Komlos Complexity}
We begin by formally defining the Komols complexity and by giving some general results about for problems with low Komols complexity. The Komols complexity of a sequence set is defined similarly to it Rademacher complexity.

\begin{definition}
Let $\F$ be a family of functions $f:\X\rightarrow\R$ and $\sigma \in \{-1,1\}^m$ uniformly at random. 
The Rademacher complexity of $\F$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
R_m(\F) =  \frac{1}{m}\E_{\sigma} \sup_{f \in \F}  \left| \sum_{i=1}^{m}\sigma_i f(x_i) \right|
\]
\end{definition}
%
In other words, the Radamacher complexity of a family function $\F$ measures how much a function in the family can skew the values when weighted with random signs. The Komlos complexity measures the ability of the family function to skew an adversarial set of signs.

\begin{definition}
Let $\F$ be a family of functions $f:\X\rightarrow\R$ and $\sigma \in \{-1,1\}^m$. 
The Komlos complexity of $\F$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
K_m(\F) =  \frac{1}{m}\min_{\sigma} \sup_{f \in \F}  \left|\sum_{i=1}^{m}\sigma_i f(x_i)\right|
\]
The Komlos Complexity $K_m(\F)$ without a reference set $\{x_1,\ldots,x_m\}$ is the upper bound on any such set.
\end{definition}

\noindent It well known that $R_m$ is a good measure for generalization when data is chosen uniformly at random (or drawn from an unknown distribution). See \cite{Bartlett:2003:RGC:944919.944944} for other notions of complexity and relationships between them. We claim that $K_m$ is the measure more suitable for creating coresets, and for creating mergeable sketches.

\el{Add a much longer explanation about the connection to discrepancy and the ideas behind this construction} The name Komlos Complexity is due to the Komlos Conjecture. It claims that $\min_\sigma \max_f \|\sum_i \sigma_i f_i\|_\infty = O(1)$ for unit vectors $f$. The correctness of the Komlos Conjecture is still a fundamental open problem in the study of discrepancy theory.  The definition is useful, however, regardless of the correctness or provability of the conjecture.  
It is obvious that  $K_m(\F) \le R_m(\F)$ for any family $\F$. We show later that $K_m(\F) = o(R_m(\F))$ for a wide range of interesting problems in machine learning.

 \zk{I actually am still having problems with justifying the name Komlos complexity. It might as well be called Banaszczyk's complexity or simply discrepancy complexity. In particular if you could prove Komlos' conjecture it wouldn't affect any of our results, nor simplify our proofs.. Either way, below is my suggested motivation for the definition}
 \el{You are right about the name. In fact, I've been thinking about it myself. The discrepancy we get is actually much lower than what Komlos conjecture says... So, I am happy with with calling it simple \it{the discrepancy.}}
To understand the motivation of the the Komlos Complexity, consider the following informal explanation of Radamacher Complexity applied to ML problems. In this ML problem there exists a distribution over $(x,y)$ pairs where we aim to predict $y$ given $x$. Our goal is to find a predictor from a family of prediction functions that suffers the least loss. A predictor maps to a loss value on each $x,y$ pair so we can discuss the problem of finding a function $f$ from a family $\F$ minimizing $\E_{x,y} f(x,y)$ where $f$ is the loss associated with the corresponding predictor on the $x,y$ pairs. Having a low Radamacher complexity means that if we sample $m$ pairs, split them into two uniformly at random and find the function $f$ minimizing the loss over the first half, the loss over the second half is bounded. This translates to a generalization bound stating that the function $f$ minimizing the loss over a uniform random sample of $(x,y)$ pairs, will perform well on the entire distribution. In other words, the Radamacher complexity gives a guarantee for the distribution loss of the minimizer of a finite set of pairs chosen uniformly at random.

Coming back to Komlos Complexity, having the ability to choose the signs in a non-uniform way lets us choose a subset of pairs $(x,y)$ such that the minimizer $f$ over the subset is guaranteed to have (roughly) the same performance on the entire set. This allows us to obtain a guarantee for the distribution loss of a minimizer of a finite set of pairs chosen in an adaptive way to the family function $\F$. This set is in fact a core-set and the Komlos complexity helps us determine that possible size of a coreset. The tradeoff of considering a coreset vs.\ a random sample is that on one hand, it is clearly much simpler to obtain a random sample. On the other hand we will show in the following sections several examples for which a coreset can be significantly smaller than the random sample while maintaining the same guarantees. This will be done by showing that for a wide range of interesting problems in machine learning $K_m(\F) = o(R_m(\F))$.
\el{I like this explanation, I'm just not sure why it talks about pairs...}

@@

>>>>

\section{Introduction}
In both machine learning and in streaming and sketching problems our goal is (usually) to approximate the sums or expectations of well behaved functions.
Specifically, we need to approximate $\E_x f(x)$ or $\frac{1}{n}\sum_{i=1}^{n} f(x_i)$ for every $f\in \F$ where $\F$ is a family of functions and $x_i \in \mathcal X$ are either training examples or stream items. 
Standard generalization results show that for a large enough $n$ the average approximates the mean if the complexity of $\F$ is bounded and the samples $x_i$ are drawn from underlying distribution. We therefore focus on approximating the average, or rather, the sum $\sum_{i=1}^{n} f(x_i)$. For notational convenience we use a parameter $q \in \mathcal Q$ to index into $\F$ explicitly $F(q) = \sum_{i=1}^{n} f(x_i, q)$ . It is convenient to think about $q$ as either the model parameters or a query for the sketch.

The goal is to produce a coreset. This is a set of weights $w$ with at most $k \ll n$ non-zeros such that $\sum_{i=1}^{n}w_i f(x_i,q) := \tilde F(q)$ approximates $F(q)$.
Approximation here means that $|\tilde F(q)  - F(q)| \le \eps n$ either for all $q \in \mathcal Q$ simultaneously or for every fixed $q$ with probability at least $1-\delta$. 
There are more complicated formulations such as weak coresets which we will not touch upon in this manuscript. 
Generating a concise representation $\tilde F$ for $F$ allows one to optimize over $\tilde F$ instead of $F$ which is more efficient.  Moreover, if the resulting coresets are mergeable, this could be done on separate machines without the need for communication or assuming randomness in the partitioning.

For bounded functions $f$ uniform sampling of $O(\log(1/\delta)/\eps^2)$ combined with a union bound over $|\mathcal Q|$ always provides a valid solution using $O(\log(|\mathcal Q|)/\eps^2)$ items. 
While $|\mathcal Q|$ is often infinite it reduces to a finite (albeit usually exponentially large) set through standard epsilon net arguments. 
%
We argue that we can offer a solution for producing smaller coresets for a large class of problems in a unified manner. 
Moreover, our solution creates streaming algorithms with fully mergeable sketches. 
The size of the coreset above appears to be intimately tied to the discrepancy properties of the associated functions.

\section{Discrepancy of a Function Class}
We begin by formally defining the discrepancy complexity and by giving some general results about for problems with low discrepancy. We remind the reader the definition of the Rademacher complexity.

\begin{definition}
Let $\F$ be a family of functions $f:\X\rightarrow\R$ and $\sigma \in \{-1,1\}^m$ uniformly at random. 
The Rademacher complexity of $\F$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
R_m(\F) =  \frac{1}{m}\E_{\sigma} \sup_{f \in \F}  \left| \sum_{i=1}^{m}\sigma_i f(x_i) \right|
\]
\end{definition}
%
In other words, the Radamacher complexity of a family function $\F$ measures how much a function in the family can skew the values when weighted with random signs. The discrepancy complexity measures the ability of the family of functions to skew an adversarial set of signs.

\begin{definition}
Let $\F$ be a family of functions $f:\X\rightarrow\R$ and $\sigma \in \{-1,1\}^m$. 
The discrepancyof $\F$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
K_m(\F) =  \frac{1}{m}\min_{\sigma} \sup_{f \in \F}  \left|\sum_{i=1}^{m}\sigma_i f(x_i)\right|
\]
The discrepancy $K_m(\F)$ without a reference set $\{x_1,\ldots,x_m\}$ is the upper bound on any such set.
\end{definition}
The following is an alternative and equivalent definition:
\begin{definition}
Let $f:\X,\Q\rightarrow\R$ and $\sigma \in \{-1,1\}^m$. 
The discrepancy of $f$ w.r.t.\ $\{x_1,\ldots,x_m\} \subset \X$ is 
\[
K_m(\F) =  \frac{1}{m}\min_{\sigma} \sup_{q \in \Q}  \left|\sum_{i=1}^{m}\sigma_i f(x_i,q)\right|
\]
The Komlos Complexity $K_m(f)$ without a reference set $\{x_1,\ldots,x_m\}$ is the upper bound on any such set.
\end{definition}

\noindent It is well known that $R_m$ is a good measure for generalization when data is chosen uniformly at random (or drawn from an unknown distribution). See \cite{Bartlett:2003:RGC:944919.944944} for other notions of complexity and relationships between them. We claim that $K_m$ is the measure more suitable for creating coresets and for creating mergeable sketches.
Trivially,  $K_m \le R_m$. We show, however, that $K_m(\F) = o(R_m(\F))$ for a wide range of interesting problems in machine learning.

To understand our motivation, consider the following informal explanation of the Radamacher Complexity applied to ML problems. 
In PAC learning there exists a set of examples (with labels).
We aim to find a predictor/classifier from a given family that suffers the least loss on the set. 
Having a low Radamacher complexity means that if we

Consider optimizing over a sample of roughly half the examples at random (each w.p.\ $1/2$).
Low Rademacher complexity guaranties that twice the loss on the sample is roughly the same as the loss on the entire set.
This translates to a generalization bound. 
In other words, the Radamacher complexity gives a guarantee for the loss of corests chosen uniformly at random.

Coming back to Discrepancy. Having the ability to choose the signs algorithmically (non-uniform) lets us choose an advantageous subset of examples.
We can actively choose to minimize the induced error and guaranty to have (roughly) the same performance on the entire set. 
This set is in fact a core-set. The Discrepancy complexity helps us determine possible sizes of a coresets. 
We will show in the following sections several examples for which a coreset can be significantly smaller than the random sample while maintaining the same guarantees. This will be done by showing that for a wide range of interesting problems in machine learning $K_m(\F) = o(R_m(\F))$.

@@

\item Kernel density estimation is a popular tool in data analysis aimed to estimate a continuous distribution with a finite set of points. Among other applications, this tool is used for outlier detection \cite{schubert2014generalized}, regression \cite{fan2018local}, and clustering \cite{rinaldo2010generalized}. A thorough survey could be found in \cite{silverman2018density}. Given a set of $n$ data points $x_1,\ldots,x_n$ and a query $q$, the density estimate of $q$ is the average $\frac{1}{n} \sum_i K(x_i,q)$ of the kernel evaluations with the data points. The task of finding a coreset for kernel density estimates consists of obtaining a set of points $z_1,\ldots,z_m$ with weights $w_1,\ldots,w_m$ for some $m \ll n$ such that for any query $q$ it holds that 
$\frac{1}{n} \sum_i K(x_i,q) = \frac{1}{m} \sum_i w_i K(z_i,q) \pm \eps$. For this task, the state-of-the-art is given by \cite{DBLP:journals/corr/abs-1802-01751}, achieving $m=\sqrt{d\log(1/\eps)}/\eps$ where $d$ is the dimension of the original data points. Their result holds for Lipchitz bounded positive semi-definite kernels. The result is constructive though it is polynomial rather than (quasi-)linear in the data size. The authors give an almost matching lower bound of $\sqrt{d}/\eps$ and pose an open question for closing the gap between the bounds.

\item Logistic Regression is an extremely popular technique for classification. We are given a set of $n$ pairs $x_i,y_i$ with $x_i \in \R^d$ and $y_i \in \{-1,1\}$. The objective is to find a regressor $q \in \R^d$ minimizing the logistic loss defined as $\E_i \log(1+\exp(x_i^\top q y_i))$. A coreset for logistic regression is a set of $m \ll n$ pairs $z_i, v_i$ in $\R^d, \{-1,1\}$ correspondingly along with weights $w_i$, with the property that for any regressor $q$, either $\E_i \log(1+\exp(x_i^\top q y_i)) = \E_i \log(1+\exp(z_i^\top q v_i)) \pm \eps$ or $\E_i \log(1+\exp(x_i^\top q y_i)) = \E_i \log(1+\exp(z_i^\top q v_i)) (1 \pm \eps)$. A recent paper \cite{DBLP:journals/corr/abs-1805-08571} provides a coreset with a multiplicative guarantee that is based on an average sensitivity property of the dataset. They provide a lower bound for the size of a multiplicative coreset showing that in general it is not possible to achieve $m \ll n$. The coreset they build is of cardinality $m \approx \mu\sqrt{nd^3}/\eps^2$ where $\mu \geq 1$ is complexity measure of the dataset.  \cite{tolochinsky2018coresets} gives a generic multiplicative coreset construction for any monotonic function with $\ell_2^2$ regularization. The dependence they get is $\tilde O(d/\eps^2)$ ignoring logarithmic factors.
